---
title: "A Principal Odor Map Unifies Diverse Tasks in Olfactory Perception"
author: "Brian K. Lee†, Emily J. Mayhew†, Benjamin Sanchez-Lengeling, Jennifer N. Wei, Wesley Qian, Kelsie Little, Matthew Andres, Britney B. Nguyen, Theresa Moloy, Jacob Yasonik, Jane Parker, Richard C. Gerkin, Joel D. Mainland, Alex Wiltschko"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    df_print: paged
    fig_height: 4.5
    fig_width: 7
    highlight: tango
    number_sections: yes
    theme: cosmo
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction to Contents

This markdown generates figures for the manuscript "A Principal Odor Map Unifies Diverse Tasks in Olfactory Perception" using the supplemental datasets made publicly available upon publication.

DOI:

# Code to Generate Figures

## Table-setting

### Load packages
```{r}
pacman::p_load(tidyverse, #data cleaning, reshaping, and plotting
               waffle, #make waffle plots
               cowplot, #stitch plots together
               ggbreak, #include axis breaks in plots
               RColorBrewer, #color palettes
               viridis, #color palettes
               svglite ) #save .svg files
```


### Read in datasets
```{r}
#Metadata for 400 molecules comprising the prospective validation dataset
S1 <- read.csv("lee_et_al_2023/data/Data S1.csv") %>%
  rename("8.atoms.or.fewer" = "X..8.atoms...25.ile.",
          "13.atoms.or.more" = "X..13.atoms...75.ile.",
          "Alkyl.4.carbon.chain" = "Alkyl")%>%
  #replace NA with empty cell
  mutate_all(~replace(., is.na(.), ""))%>%
  mutate(RedJade.Code = as.numeric(RedJade.Code))

#Panelist evaluations of 400 novel odorants
S3 <- read.csv("lee_et_al_2023/data/Data S3.csv")

#Odor attribute predictions on 400 molecules by a random forest model trained on GS/LF datasets
S4 <- read.csv("lee_et_al_2023/data/Data S4.csv")

#Odor attribute predictions on 400 molecules by a graph neural network model trained on GS/LF datasets. Final layer.
S5 <- read.csv("lee_et_al_2023/data/Data S5.csv")

#Graph neural network embeddings on 400 molecules. Penultimate layer.
S6 <- read.csv("lee_et_al_2023/data/Data S6.csv")

#Correspondence table between internal odorant identifiers and chemical structures
S7 <- read.csv("lee_et_al_2023/data/Data S7.csv")
```


## Figure 1B
The GNN was trained on a curated dataset of ~5000 semantically labeled molecules drawn from GoodScents (13) and Leffingwell (14) flavor and fragrance databases; one square represents 100 molecules.
```{r, message=F, warning=F}
#create df with database sizes
traindata <- c(GoodScents = 37.86 - 23.17, Both = 23.17, Leffingwell = 35.61 - 23.17)

#generate waffle plot
waffle(traindata, rows = 10, flip = T, legend_pos = "bottom", size = 1, reverse = T)

#remove df from environment
rm(traindata)
```


## Figure 2A-D
(A) GNN model label predictions, (B) random forest (RF) model label predictions, (C) panel mean ratings with standard error bars, and (D) individual panelist ratings, averaged over 2 replicates, for the molecule 2,3-dihydrobenzofuran-5-carboxaldehyde. In panels A-C, the top 5 ranked descriptors are in orange (GNN), purple (RF), or green (panel). Descriptors in panels A-D are ordered by panel mean ratings. Panels A, B, and D are annotated with the Pearson correlation coefficient of their data to the panel mean rating shown in panel C. Panel D includes panelist/panel correlation coefficients for the panelist that best matches the panel mean and for the panelist with the median match.
```{r, message=F, warning=F}
#normalize data by subtracting from each raw rating that subject's mean rating for the given descriptor
RATA_normalize <- S3 %>%
  group_by(SubjectCode)%>%
  summarise_at(vars(Green:Fishy), mean) %>%
  pivot_longer(Green:Fishy, names_to = "Descriptor", values_to = "SubjectMean")

RATA_normalized <-  S3 %>%
  pivot_longer(Green:Fishy, names_to = "Descriptor", values_to = "Value")%>%
  left_join(., RATA_normalize)%>%
  mutate(normValue = Value - SubjectMean)
```

```{r, message=F, warning=F}
#Calculate subject correlation to group mean

#correlation to mean ratings - removing evaluation to calculate mean
ChemCor <- RATA_normalized %>%
  select(SubjectCode, Rep, SampleIdentifier, Descriptor, normValue)

df <- ChemCor %>% filter(SubjectCode != unique(S3$SubjectCode)[1]) %>%
  group_by(SampleIdentifier, Descriptor) %>%
  summarize(GroupMean = mean(normValue)) 

df2 <- ChemCor %>% filter(SubjectCode == unique(S3$SubjectCode)[1]) %>%
  select(SampleIdentifier, Descriptor, SubjectCode, Rep, normValue) %>%
  left_join(df)

results <- df2 

for(i in 2:length(unique(S3$SubjectCode))){
  df <- ChemCor %>% filter(SubjectCode != unique(S3$SubjectCode)[i]) %>%
    group_by(SampleIdentifier, Descriptor) %>%
    summarize(GroupMean = mean(normValue)) 
  
  df2 <- ChemCor %>% filter(SubjectCode == unique(S3$SubjectCode)[i]) %>%
    select(SampleIdentifier, Descriptor, SubjectCode, Rep, normValue) %>%
    left_join(df)
  
  results <- bind_rows(results, df2)
}
ChemCor2 <- results %>%
  group_by(SubjectCode, Rep, SampleIdentifier)%>%
  summarise(corP = cor(normValue, GroupMean))

#clean up environment
rm(df, df2, ChemCor, RATA_normalize)
```

```{r, message=F, warning=F}
#pick an example molecule
sampleID <- 380 

#now using color brewer palette 2
palette <- brewer.pal(n=4, "Set2")
#order: green - people, orange - gnn, purple - rf, pink - other

#calculate correlation of subject ratings to group mean for chosen molecule
cor_subj <- ChemCor2 %>%
  filter(SampleIdentifier == sampleID) %>%
  group_by(SubjectCode) %>%
  summarise(corP = mean(corP)) %>%
  mutate(SubjectCode = reorder(SubjectCode, corP, FUN = mean))
```


```{r, message=F, warning=F}
#make new df with mean ratings, GNN predictions, and RF predictions

#summarize to mean values
RATA.meanVal <- S3 %>% 
  select(-"StartTime.UTC.", -Intensity, -Pleasantness, -NoOdor )%>%
  pivot_longer(Green:Fishy, names_to = "Descriptor", values_to = "Value") %>%
  group_by(SampleIdentifier, Descriptor) %>%
  summarise(MeanRating = mean(Value),
            num = n(),
            stdev = sd(Value),
            se = stdev/sqrt(num))

#make dataframes long
gnnP_long <- S5 %>%
  pivot_longer(-RedJade.Code, names_to = "Descriptor", values_to = "GNNprediction")%>%
  rename(SampleIdentifier = RedJade.Code)%>%
  mutate(SampleIdentifier = as.numeric(SampleIdentifier))

rf_long <- S4 %>%
  pivot_longer(-RedJade.Code, names_to = "Descriptor", values_to = "RFprediction")%>%
  rename(SampleIdentifier = RedJade.Code)%>%
  mutate(SampleIdentifier = as.numeric(SampleIdentifier))


#merge gnn, rf, and rata ratings
ratings.gnn <- left_join(RATA.meanVal, gnnP_long)%>%
  left_join(., rf_long) %>%
  group_by(SampleIdentifier)%>%
  mutate(gnnRank = rank(GNNprediction),
         rfRank = rank(RFprediction),
         rataRank = rank(MeanRating),
         gnnTop5 = ifelse(gnnRank > 50, "yes", "no"),
         rfTop5 = ifelse(rfRank > 50, "yes", "no"),
         rataTop5 = ifelse(rataRank > 50, "yes", "no"))

#clean up environment
rm(RATA.meanVal, gnnP_long, rf_long)
```


```{r, message=F, warning=F}
#make RF part of the figure
rffig <-   ratings.gnn %>%
  mutate(rfTop5 = factor(rfTop5,
                          levels = c("yes", "no")))%>%
  filter(SampleIdentifier == sampleID) %>%
  ggplot( aes( reorder(Descriptor, -MeanRating), RFprediction, fill = rfTop5))+
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c(palette[3], "gray"),
                    labels = c("Top 5 predicted labels", "All other labels"))+
  theme_classic()+
  scale_y_continuous(expand = c(0,0)) +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        plot.margin = unit(c(0, 0, 0, 0), "cm"),
        legend.position = c(0.9, 0.8),
        legend.title = element_blank())+
  ylim(0, 1)+
  ylab("RF \np(label applies)")
```

```{r, message=F, warning=F}
#make GNN part of the figure
gnnfig <-   ratings.gnn %>%
  mutate(gnnTop5 = factor(gnnTop5,
                          levels = c("yes", "no")))%>%
  filter(SampleIdentifier == sampleID) %>%
  ggplot( aes( reorder(Descriptor, -MeanRating), GNNprediction, fill = gnnTop5))+
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c(palette[2], "gray"),
                    labels = c("Top 5 predicted labels", "All other labels"))+
  theme_classic()+
  scale_y_continuous(expand = c(0,0)) +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        plot.margin = unit(c(0, 0, 0, 0), "cm"),
        legend.position = c(0.9, 0.8),
        legend.title = element_blank())+
  ylim(0, 1)+
  ylab("GNN \np(label applies)")
```

```{r, message=F, warning=F}
#make rata part of the figure
ratafig <- ratings.gnn %>%
  mutate(rataTop5 = factor(rataTop5,
                           levels = c("yes", "no")))%>%
  filter(SampleIdentifier == sampleID) %>%
  ggplot( aes( reorder(Descriptor, -MeanRating), MeanRating, fill = rataTop5))+
  geom_bar(stat = "identity")+
  geom_errorbar( aes(ymin = MeanRating - se, ymax = MeanRating + se))+
  scale_fill_manual(values = c(palette[1], "gray"),
                    labels = c("Top 5 rated labels", "All other labels"))+
  theme_classic()+
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        plot.margin = unit(c(0, 0, 0, 0), "cm"),
        legend.position = c(0.88, 0.8),
        legend.title = element_blank())+
  ylab("Panel mean rating")
```

```{r, message=F, warning=F}
#raster plot of raw ratings

#stack descriptors for easy plotting
RATA.long <- S3 %>%
  select(-1, -Intensity, -Pleasantness, -NoOdor)%>%
  pivot_longer(Green:Fishy, names_to = "Descriptor", values_to = "Value")

#raw data plot
raster <- RATA.long %>%
  filter(SampleIdentifier == sampleID) %>%
  group_by(Descriptor)%>%
  mutate(MeanRating = mean(Value))%>%
  rename(Rating = Value)%>%
  left_join(., cor_subj)%>%
  ggplot( aes(reorder(SubjectCode, corP), reorder(Descriptor, -MeanRating), fill = Rating))+
  geom_raster() +
  coord_flip()+
  xlab("Panelists")+
  scale_fill_distiller(type = "seq", direction = 1, palette = 2) +
  theme(legend.position = c(0.95, 0.55),
        legend.key.height = unit(0.45, 'cm'),
        legend.text = element_text(size = 8),
        axis.text.x = element_text(size=9, angle = 60, hjust = 1),
        axis.title.x = element_blank(), 
        axis.text.y = element_text(size = 6))
```


```{r, message=F, warning=F}
#put subplots together
plot_grid( gnnfig, NULL, rffig, NULL, ratafig, NULL, raster, ncol = 1, align = c("hv"), 
           rel_heights = c(.6, -0.22, 0.6, -0.22, 0.6, -0.22, 0.7))

#clean up 
rm(rffig, gnnfig, ratafig, raster)
```


## Figure 2F
(F) Difference in the median correlation to the panel mean relative to the median human subject’s correlation to the panel mean for models trained using k-nearest neighbor (KNN) and RF, trained on cFPs or Mordred features, and the GNN model. Only the GNN model has a median correlation to the panel mean that is higher than that of the median panelist. 
```{r, message=F, warning=F}
#plot just RF and GNN relative to human baseline

#make df with median correlation coefficients to panel mean (calculations done in other script)
#these values from Brian, updated as of 3/24/2023
OTT <- tibble(Human = 0.207,
                  GNN = 0.242,
                  RF_cFP = 0.176,
                  RF_Mordred = 0.187,
                  KNN_cFP = 0.127,
                  KNN_Mordred = 0.148,
                  GNN_shuffled = -0.045803) %>% #no updated value given
  pivot_longer(Human:GNN_shuffled, names_to = "Model", values_to = "MedianCorrelation") %>%
  mutate(PerformanceDelta = MedianCorrelation - 0.210246) #compare all corr to human baseline 
  
#make barplot with broken axis
OTT %>%
  filter(! Model %in% c("Human", "RF", "GNN_retrained"))%>%
  ggplot( aes(fct_reorder(Model, PerformanceDelta, mean), PerformanceDelta, fill = Model))+
  geom_col()+
  geom_hline(yintercept = 0)+
  theme_classic()+
  scale_fill_manual(values = c( "#fc8d62", "#b3b3b3", 
                                "#E62C8B", "#E62C8B","#8da0cb", "#8da0cb"))+
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        text = element_text(size = 8),
        axis.text.x = element_text(angle = 30, hjust = 1))+
  ylim(-0.3, 0.05)+
  scale_y_break(c(-0.2, -0.1))+
  ylab("Difference in Median Correlation \nRelative to Human Baseline")

#remove df from environment
rm(OTT)
```


## Figure 3A
(A) Correlation of GNN (in orange) and RF (in purple) model predictions and panelist ratings (in gray) to the panel mean for each of the 55 odor labels. 
```{r, message=F, warning=F}
#calculate normalized mean RATA ratings
RATA_normMeans <-  RATA_normalized %>%
  group_by(SampleIdentifier, Descriptor)%>%
  summarise(meanRating = mean(normValue))%>%
  rename(RedJade.Code = SampleIdentifier)

#normalize predictions
GNN_normalize <- S5 %>%
  summarise_at(vars(Green:Fishy), mean) %>%
  pivot_longer(Green:Fishy, names_to = "Descriptor", values_to = "Mean")

RF_normalize <- S4 %>%
  summarise_at(vars(Green:Fishy), mean) %>%
  pivot_longer(Green:Fishy, names_to = "Descriptor", values_to = "Mean")

#make long, normalize
gnnNorm_long <- S5 %>%
  pivot_longer(-RedJade.Code, names_to = "Descriptor", values_to = "GNNprediction")%>%
  left_join(., GNN_normalize)%>%
  mutate(nGNN = GNNprediction - Mean)%>%
  mutate(RedJade.Code = as.numeric(RedJade.Code))%>%
  select(RedJade.Code, Descriptor, nGNN)
  
rfNorm_long <- S4 %>%
  pivot_longer(-RedJade.Code, names_to = "Descriptor", values_to = "RFprediction")%>%
  left_join(., RF_normalize)%>%
  mutate(nRF = RFprediction - Mean)%>%
  mutate(RedJade.Code = as.numeric(RedJade.Code))%>%
  select(RedJade.Code, Descriptor, nRF)

# #S1 is not up-to-date on dropped molecules
# odor_key_resub <- read.csv("odor_key_resub.csv")

#calculate mean model correlations
ModelCor <- RATA_normMeans %>%
  #left_join(., odor_key_resub)%>%
  left_join(., S1) %>%
  left_join(., gnnNorm_long) %>%
  left_join(., rfNorm_long) %>%
  filter(Disqualification.reason == "") %>%
  drop_na()%>%
  group_by(Descriptor)%>%
  summarise(corGNN = cor(nGNN, meanRating),
            corRF = cor(nRF, meanRating))

 #calculate mean subject correlations, merge with model correlations 
LabelCor <- results %>%
  rename(RedJade.Code = SampleIdentifier)%>%
  #left_join(., odor_key_resub)%>%
  left_join(., S1) %>%
  filter(Disqualification.reason == "") %>%
  drop_na()%>%
  group_by(SubjectCode, Rep, Descriptor)%>%
  summarise(corP = cor(normValue, GroupMean))%>%
  drop_na()%>%
  left_join(ModelCor)


##### Label figure ########

min(LabelCor$corP)
max(LabelCor$corP)

LabelCor %>%
  ggplot( aes(corP, reorder(Descriptor, corGNN))) +
  geom_boxplot(outlier.alpha = 0, color = "lightgray")+
  geom_point(alpha = 0.05)+ 
  geom_point( aes(corGNN, Descriptor), color = palette[2], size = 2.5)+
  geom_point( aes(corRF, Descriptor), color = palette[3], shape = 18, size = 2.5)+ #matches barcode plot orange
  theme_classic()+
  xlim(-0.13, 0.92)+
  xlab("Correlation to Panel Mean")+
  theme(axis.title.y = element_blank(),
        axis.text.y = element_text(size=8))

```

```{r}
#pull some accompanying stats
df1 <- LabelCor %>%
  group_by(Descriptor) %>%
  summarise(GNNinDist = case_when(corGNN > min(corP) ~ "yes",
                               TRUE ~ "no"),
         RFinDist = case_when(corRF > min(corP) ~ "yes",
                              TRUE ~ "no"),
         GNNbeatsMedian = case_when(corGNN > median(corP) ~ "yes",
                                    TRUE ~ "no"),
         RFbeatsMedian = case_when(corRF > median(corP) ~ "yes",
                              TRUE ~ "no"),
         GNNbeatsRF = case_when(corGNN > corRF ~ "yes",
                                TRUE ~ "no"))
#GNN  
#GNN below distribution
df1%>%
  unique()%>%
  filter(GNNinDist == "no")%>%
  select(Descriptor) # just Musk

#GNN above median panelist
df1%>%
  unique()%>%
  filter(GNNbeatsMedian == "yes")%>%
  select(Descriptor) #30 labels
30/55 #55%

#RF
#RF below distribution
df1%>%
  unique()%>%
  filter(RFinDist == "no")%>%
  select(Descriptor) # 2 - fermented, sharp

#RF above median panelist
df1%>%
  unique()%>%
  filter(RFbeatsMedian == "yes")%>%
  select(Descriptor) #16 labels
16/55 #29%


#RF v GNN
df1%>%
  unique()%>%
  filter(GNNbeatsRF == "yes")%>%
  select(Descriptor) #46 labels
46/55 #84%


## RF vs GNN t-test for label performance
label.ttest <- LabelCor %>%
  group_by(Descriptor)%>%
  summarise(GNNmean = mean(corGNN),
            RFmean = mean(corRF))

t.test(x=label.ttest$GNNmean, y=label.ttest$RFmean, paired = TRUE)
#p-value = 3.349e-07

```


## Figure 3C
(C) Mean correlation of GNN (in orange) and RF (in purple) model predictions and panelist ratings (in gray) to the panel mean for molecules belonging to 12 common chemical classes.

```{r, message=F, warning=F}

#calculate mean model correlations, group by molecule
ModelCor <- RATA_normMeans %>%
  #left_join(., odor_key_resub)%>%
  left_join(., S1) %>%
  left_join(., gnnNorm_long) %>%
  left_join(., rfNorm_long) %>%
  drop_na()%>%
  group_by(RedJade.Code, Disqualification.reason)%>%
  summarise(corGNN = cor(nGNN, meanRating),
            corRF = cor(nRF, meanRating))

#S1 also needs to be updated to include chemical class categorizations
corr_class <- S1 %>% 
  select(RedJade.Code, Sulfur:"13.atoms.or.more") %>%
  pivot_longer(Sulfur:"13.atoms.or.more", names_to = "ChemicalClass", values_to = "Member")%>%
  filter(Member == "TRUE")%>%
  group_by(ChemicalClass)%>%
  select(RedJade.Code, ChemicalClass)


ChemCor3 <- ChemCor2 %>%
  rename(RedJade.Code = SampleIdentifier)%>%
  left_join(., ModelCor)%>%
  #left_join(., S1) %>%
  filter(Disqualification.reason == "") %>%
  left_join(., corr_class) %>%
  filter(!is.na(ChemicalClass))%>% #some molecules don't fit any class
  mutate(ChemicalClass = str_replace_all(ChemicalClass, "\\.", " "))%>%
  mutate(ChemicalClass = str_replace_all(ChemicalClass, "4 carbon", "4-carbon"))


Disagg_chem_models <- ChemCor3 %>%
  group_by(ChemicalClass)%>%
  summarise(GNNmean = mean(corGNN, na.rm = T),
         RFmean = mean(corRF, na.rm = T))%>%
  mutate(ChemicalClass = fct_reorder(ChemicalClass, GNNmean))

Disagg_chem <- ChemCor3 %>%
  group_by(SubjectCode, ChemicalClass) %>%
  summarise(corPsubjmean = mean(corP, na.rm = T)) %>%
  left_join(Disagg_chem_models) %>%
  mutate(ChemicalClass = fct_reorder(ChemicalClass, GNNmean))
 

#plot
min(Disagg_chem$corPsubjmean)
max(Disagg_chem$corPsubjmean)

ggplot(Disagg_chem, aes(corPsubjmean, ChemicalClass)) +
  geom_boxplot(outlier.alpha = 0, color = "lightgray")+
  geom_point(alpha = 0.05)+ 
  geom_point(data = Disagg_chem_models, aes(GNNmean, ChemicalClass), color = palette[2], size = 2.5)+
  geom_point(data = Disagg_chem_models, aes(RFmean, ChemicalClass), color = palette[3], shape = 18, size = 2.5)+ #matches barcode plot orange
  theme_classic()+
  xlim(-0.15, 0.92)+
  xlab("Correlation to Panel Mean")+
  theme(axis.title.y = element_blank(),
        axis.text.y = element_text(size=8))


#stats on chemical class performance
min(Disagg_chem$GNNmean) #0.16
max(Disagg_chem$GNNmean) #0.52


#clean up environment
rm(corr_class, Disagg_chem, Disagg_chem_models, GNN_normalize, gnnNorm_long, RF_normalize, rfNorm_long)
```


```{r}
#GNN vs median panelist // RF vs median panelist // GNN vs RF

#filter model_cor to non-dropped molecules
ModelCor_thinned <- ModelCor %>%
  filter(Disqualification.reason == "")

cor_medianPanelist <- ChemCor2 %>%
  group_by(SampleIdentifier) %>%
  summarize(MedianR = median(corP))%>%
  rename(RedJade.Code = SampleIdentifier)

OTT_test <- ModelCor_thinned %>%
  left_join(cor_medianPanelist) %>%
  mutate(GNN.beats.Median = case_when(corGNN > MedianR ~ "yes",
                                TRUE ~ "no"),
         RF.beats.Median = case_when(corRF > MedianR ~ "yes",
                                TRUE ~ "no"))

#calculate percent of cases in which GNN or RF come closer to panel mean than median panelist
OTT_test %>%
  group_by(GNN.beats.Median)%>%
  tally()
175/323 #54.2% 

OTT_test %>%  
  group_by(RF.beats.Median)%>%
  tally()
141/323 #43.7% 

#conduct 2-sided paired t-test comparing each pair of GNN, RF, and median panelist correlation to panel mean (per molecule)
t.test(OTT_test$corGNN, OTT_test$MedianR, alternative = "two.sided", paired = TRUE) #p-value = 0.519
t.test(OTT_test$corRF, OTT_test$MedianR, alternative = "two.sided", paired = TRUE) #p-value = 5.22e-05
t.test(OTT_test$corRF, OTT_test$corGNN, alternative = "two.sided", paired = TRUE) #p-value = 2.253e-09

mean(cor_medianPanelist$MedianR)
```




## Figure 3D
(D) Categorization of gas chromatography-olfactometry quality control results for 50 validation set stimuli. 
```{r, message=F, warning=F}
GCO <- c("Odor caused by nominal compound" = 11, 
         "Minor impact on odor by contaminants" = 15,
         "Cause of odor cannot be determined" = 4, 
         "Major impact on odor by contaminants" = 5, 
         "Odor not caused by nominal compound" = 15)

waffle(GCO, rows = 5, flip = F, legend_pos = "right", size = 1, glyph_size = 1, reverse = F,
       colors = c("#2166ac", "#67a9cf", "#cccccc", "#f7941d", "#b2182b"))

```


## Figure S9
Panel mean (n≥15 subjects) test-retest correlation (R) for the 55 descriptors in the lexicon applied to the 400 novel odorants in the prospective validation set. Each dot represents 1 molecule. Descriptors are ordered by descending correlation. 
```{r, message=F, warning=F}
#plot mean ratings from rep1 vs rep2
meansTRT <- S3 %>%
  pivot_longer(NoOdor:Fishy, names_to = "Descriptor", values_to = "Value") %>%
  group_by(SampleIdentifier, Descriptor, Rep) %>%
  summarise(GroupMean = mean(Value, na.rm = T)) %>%
  pivot_wider(names_from = "Rep", values_from = "GroupMean", names_prefix = "Rep")

##### mean TRT for all Descriptors, faceted by Descriptor, sorted by R #####
meansTRT %>%
  group_by(Descriptor)%>%
  mutate(R =  cor(Rep1, Rep2),
         Rlabel = paste0("R=", round(cor(Rep1, Rep2),2)))%>%
  filter(Descriptor != "NoOdor")%>%
  ggplot( aes(Rep1, Rep2, label= Rlabel)) +
  geom_point( alpha=0.6)+
  geom_smooth(method = "lm", color = "royalblue")+
  geom_text(color = "blue",
            x=1.5, y=4, size = 3)+
  facet_wrap(~fct_reorder(Descriptor, R, .desc = T))+
  scale_x_continuous(limits = c(0,5), expand = c(0,0))+
  scale_y_continuous(limits = c(0,5), expand = c(0,0))+
  theme_classic()+
  #scale_color_manual(values = c("steelblue1", "maroon", "gold2"))+
  #scale_shape_manual(values = c(17, 16, 15))+
  xlab("Panel Mean - 1st Replicate")+
  ylab("Panel Mean - 2nd Replicate")+
  theme(legend.position = "none")

```


## Figure S10 
Panel mean (n=49) test-retest correlation (R) for the 19 descriptors in the DREAM olfaction challenge dataset (6). Each dot represents 1 molecule. Descriptors are ordered by descending correlation. 
```{r, message=F, warning=F}
#read in DREAM data from Keller et al 2016 DOI: 10.1126/science.aal2014
DREAM <- read.table("lee_et_al_2023/data/DREAMTrainSet.txt", fill = T, header = T, sep = "\t", comment.char = "")

#remove replicate from odor name - redundant with replicate column
DREAM$Odor <- gsub("\\(replicate)", "", DREAM$Odor)
#remove spaces
DREAM$Odor <- str_trim(DREAM$Odor)

#find repeated odors
rep.odors <- unique(DREAM[DREAM$Replicate == "replicate ", "Odor"])

#calculate mean values
DREAMmeans <- DREAM %>%
  filter(Odor %in% rep.odors) %>%
  pivot_longer(INTENSITY.STRENGTH:CHEMICAL, names_to = "Descriptor", values_to = "Rating")%>%
  mutate(Rep = case_when(Replicate == "" ~ "Rep1",
                         TRUE ~ "Rep2"))%>%
  select(Odor, Intensity, Rep, Descriptor, subject.., Rating) %>%
  group_by(Odor, Intensity, Descriptor, Rep)%>%
  summarise(meanRating = mean(Rating, na.rm = T))%>%
  pivot_wider(names_from = "Rep", values_from = "meanRating")

#DREAM mean TRT for all Descriptors, faceted by Descriptor, sorted by R
DREAMmeans %>%
  filter(!Descriptor %in% c("INTENSITY.STRENGTH", "VALENCE.PLEASANTNESS"))%>%
  mutate(Descriptor = str_replace(Descriptor, ".URINOUS", replacement = ""))%>%
  group_by(Descriptor)%>%
  mutate(R = cor(Rep1, Rep2),
         Rlabel =  paste0("R=", round(cor(Rep1, Rep2),2)))%>%
  ggplot( aes(Rep1, Rep2, label= Rlabel)) +
  geom_point( alpha=0.6)+
  geom_smooth(method = "lm", color = "royalblue")+
  geom_text(color = "blue",
            x=24, y=85, size = 3)+
  facet_wrap(~fct_reorder(Descriptor, R, .desc = T), ncol = 8)+
  scale_x_continuous(limits = c(0,100), expand = c(0,0))+
  scale_y_continuous(limits = c(0,100), expand = c(0,0))+
  theme_classic()+
  xlab("Panel Mean - 1st Replicate")+
  ylab("Panel Mean - 2nd Replicate")+
  theme(legend.position = "none")
```



## Figure S11
Panel mean test-retest correlation for the 19 descriptors in the DREAM olfaction challenge (left) (6) and for the 55 descriptors in the present study (right). Each dot represents test-retest correlation for one odor descriptor. 
```{r, message=F, warning=F}
#summarise monell TRT, include intensity & pleasantness
Monell.TRT <- S3 %>%
  pivot_longer(Intensity:Fishy, names_to = "Descriptor", values_to = "Value") %>%
  group_by(SampleIdentifier, Descriptor, Rep) %>%
  summarise(GroupMean = mean(Value, na.rm = T)) %>%
  pivot_wider(names_from = "Rep", values_from = "GroupMean", names_prefix = "Rep")%>%
  drop_na()%>%
  group_by(Descriptor)%>%
  summarise(R = cor(Rep1, Rep2))%>%
  mutate(Cohort = "Monell")

#summarise DREAM TRT
DREAM.TRT <- DREAMmeans %>%
  group_by(Descriptor)%>%
  summarise(R = cor(Rep1, Rep2))%>%
  mutate(Cohort = "DREAM")

#calculate mean of TRT distribution
mean(Monell.TRT$R)
mean(DREAM.TRT$R)

# violin plot comparing mean TRT btw DREAM and Monell cohorts
Monell.TRT %>%
  bind_rows(., DREAM.TRT)%>%
  ggplot( ) +
  geom_violin( aes(x= Cohort, y=R), alpha=0.6, draw_quantiles = c(0.5))+
  geom_dotplot(aes(x = Cohort, y=R), binaxis = "y", binwidth = 0.025, stackdir = "center", alpha=0.6)+
  ylab("Panel mean test-retest correlation")+
  scale_y_continuous(limits = c(0,1), breaks = seq(0,1, by=0.1))+
  theme_classic()+
  theme(legend.position = "none",
        axis.title.x = element_blank()  )

#clean up environment
rm(DREAM, DREAM.TRT, DREAMmeans, meansTRT, Monell.TRT)
```

## Figure S14
For the molecules (A-D) 1-cyclopropylethanol, (E-H) 3,3-dimethylcyclohexan-1-one, (I-L) 2-Methylquinolin-8-ol, (M-P) 2-Phenoxyethylamine, (A, E, I, M) GNN model label predictions, (B, F, J, N) random forest (RF) model label predictions, (C, G, K, O) panel mean ratings with standard error bars, and (D, H, L, P) individual panelist ratings, averaged over 2 replicates, for the molecule 2,3-dihydrobenzofuran-5-carboxaldehyde. The top 5 ranked descriptors for each molecule are in orange (GNN), purple (RF), or green (panel). Descriptors in panels are ordered by panel mean ratings for the given molecule. Panels are annotated with the Pearson correlation coefficient of their data to the panel mean rating. Panels D, H, L, and P include panelist/panel correlation coefficients for the panelist that best matches the panel mean and for the panelist with the median match.
```{r, message=F, warning=F}

#Make additional versions for supplemental

#make list to receive plots
plot_list <- list()

#choose GCO clean molecules
sampleIDs <- c(6, 153, 161, 251)

#now using color brewer palette 2
palette <- brewer.pal(n=4, "Set2")
#order: green - people, orange - gnn, purple - rf, pink - other

#loop through 4 new GCO clean molecules
for(i in 1:length(sampleIDs)){
sampleID <- sampleIDs[i]
#calc cor
cor_subj <- ChemCor2 %>%
  filter(SampleIdentifier == sampleID) %>%
  group_by(SubjectCode) %>%
  summarise(corP = mean(corP)) %>%
  mutate(SubjectCode = reorder(SubjectCode, corP, FUN = mean))

#make RF part of the figure
rffig <-   ratings.gnn %>%
  mutate(rfTop5 = factor(rfTop5,
                         levels = c("yes", "no")))%>%
  filter(SampleIdentifier == sampleID) %>%
  ggplot( aes( reorder(Descriptor, -MeanRating), RFprediction, fill = rfTop5))+
  geom_bar(stat = "identity") +
  #geom_text(data = ratings.gnn %>% filter(SampleIdentifier ==6 & gnnTop5 == "yes"), 
  #          aes(y = GNNprediction-0.002, label = Descriptor), color = "black", hjust = 0, nudge_x = 1)+
  scale_fill_manual(values = c(palette[3], "gray"),
                    labels = c("Top 5 predicted labels", "All other labels"))+
  theme_classic()+
  scale_y_continuous(expand = c(0,0)) +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        plot.margin = unit(c(0, 0, 0, 0), "cm"),
        legend.position = c(0.9, 0.8),
        legend.title = element_blank())+
  ylim(0, 1)+
  ylab("RF \np(label applies)")


#make GNN part of the figure
gnnfig <-   ratings.gnn %>%
  mutate(gnnTop5 = factor(gnnTop5,
                          levels = c("yes", "no")))%>%
  filter(SampleIdentifier == sampleID) %>%
  ggplot( aes( reorder(Descriptor, -MeanRating), GNNprediction, fill = gnnTop5))+
  geom_bar(stat = "identity") +
  #geom_text(data = ratings.gnn %>% filter(SampleIdentifier ==6 & gnnTop5 == "yes"), 
  #          aes(y = GNNprediction-0.002, label = Descriptor), color = "black", hjust = 0, nudge_x = 1)+
  scale_fill_manual(values = c(palette[2], "gray"),
                    labels = c("Top 5 predicted labels", "All other labels"))+
  theme_classic()+
  scale_y_continuous(expand = c(0,0)) +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        plot.margin = unit(c(0, 0, 0, 0), "cm"),
        legend.position = c(0.9, 0.8),
        legend.title = element_blank())+
  ylim(0, 1)+
  ylab("GNN \np(label applies)")

#make rata part of the figure
ratafig <- ratings.gnn %>%
  mutate(rataTop5 = factor(rataTop5,
                           levels = c("yes", "no")))%>%
  filter(SampleIdentifier == sampleID) %>%
  ggplot( aes( reorder(Descriptor, -MeanRating), MeanRating, fill = rataTop5))+
  geom_bar(stat = "identity")+
  geom_errorbar( aes(ymin = MeanRating - se, ymax = MeanRating + se))+
  #geom_text(data = ratings.gnn %>% filter(SampleIdentifier ==6 & rataTop5 == "yes"), 
  #          aes(y = MeanRating, label = Descriptor), color = "black", hjust = 0, nudge_x = 1)+
  #scale_y_reverse()+
  scale_fill_manual(values = c(palette[1], "gray"),
                    labels = c("Top 5 rated labels", "All other labels"))+
  theme_classic()+
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        plot.margin = unit(c(0, 0, 0, 0), "cm"),
        legend.position = c(0.88, 0.8),
        legend.title = element_blank())+
  ylab("Panel mean rating")

#raw data plot
raster <- RATA.long %>%
  filter(SampleIdentifier == sampleID) %>%
  group_by(Descriptor)%>%
  mutate(MeanRating = mean(Value))%>%
  rename(Rating = Value)%>%
  left_join(., cor_subj)%>%
  ggplot( aes(reorder(SubjectCode, corP), reorder(Descriptor, -MeanRating), fill = Rating))+
  geom_raster() +
  coord_flip()+
  xlab("Panelists")+
  scale_fill_distiller(type = "seq", direction = 1, palette = 2) +
  theme(legend.position = c(0.95, 0.55),
        legend.key.height = unit(0.45, 'cm'),
        legend.text = element_text(size = 8),
        axis.text.x = element_text(size=9, angle = 60, hjust = 1),
        axis.title.x = element_blank(), 
        axis.text.y = element_text(size = 6))

#put subplots together
plot_list[[i]] <- plot_grid( gnnfig, NULL, rffig, NULL, ratafig, NULL, raster, ncol = 1, align = c("hv"), 
           rel_heights = c(.6, -0.22, 0.6, -0.22, 0.6, -0.22, 0.7))
}

plot_list
```

```{r, warning=F}
#clean up environment
rm(ChemCor2, cor_subj, gnnfig, plot_list, raster, RATA_normalized, RATA.long, ratafig, results, rffig, sampleID, sampleIDs)
```


## Figure S17
Label frequency among contaminated molecules and full test set. Plot shows percent of molecules dropped due to odorous contamination (“QC-dropped”, n = 19) with a given label (mean panel rating > 1) in orange and percent of molecules in the full test set (n = 399) with a given label (mean panel rating > 1) in blue. Odor labels such as “fishy” and “garlic” are much more common among contaminated molecules than in the full test set.
```{r, message=F, warning=F}
#filter RATA data to just labels with panel mean rating > 1
panel_df <- S3 %>%
  pivot_longer(Green:Fishy, names_to = "Descriptor", values_to = "Rating")%>%
  group_by(SampleIdentifier, Descriptor)%>%
  summarise(meanRating = mean(Rating)) %>%
  filter(meanRating > 1)%>% 
  group_by(Descriptor)%>%
  tally()%>%
  mutate(Type = "Full Test Set Panel Ratings")

#RATA profile for molecules dropped due to confirmed contaminant (verdict B or C)
drop_df <- S3 %>%
  left_join(S1, by = c("SampleIdentifier" = "RedJade.Code")) %>% 
  filter(GCO.result != "")%>% #in QC set
  filter(Disqualification.reason != "") %>% #dropped because of QC results
  # group_by(SampleName)%>% #19 molecules dropped due to contamination
  # tally()
  select(SampleIdentifier, Green:Fishy)%>%
  pivot_longer(Green:Fishy, names_to = "Descriptor", values_to = "Rating")%>%
  group_by(SampleIdentifier, Descriptor)%>%
  summarise(meanRating = mean(Rating)) %>%
  filter(meanRating > 1)%>% 
  group_by(Descriptor)%>%
  tally()%>%
  mutate(Type = "QC-dropped Molecule Ratings")

#set order of descriptors
most_contaminated <- drop_df %>%
  arrange(-n)%>%
  select(Descriptor)

use_order <- panel_df %>%
  filter(! Descriptor %in% most_contaminated$Descriptor) %>%
  arrange(-n)%>%
  select(Descriptor)

order <- most_contaminated %>%
  bind_rows(use_order)

order <- order$Descriptor

#relative frequency of labels among contaminated molecules and full test set
panel_df %>%
  bind_rows(drop_df)%>%
  #case_when was not working, resorted to nested ifelse statements
  mutate(relFreq = ifelse(Type == "Full Test Set Panel Ratings", n*100/399, n*100/19))%>%
  mutate(Type = fct_relevel(Type,
                            levels = c("QC-dropped Molecule Ratings", 
                                       "Full Test Set Panel Ratings"
                                       )))%>%
  mutate(Descriptor = fct_relevel(Descriptor, 
                                  levels = c(order)))%>%
  ggplot( aes(x = Descriptor, relFreq, fill = Type)) +
  geom_bar(stat = "identity", position = position_dodge2(reverse = F))+
  scale_fill_manual(values = c("orangered", "skyblue"))+
  theme_classic()+
  #scale_x_discrete(breaks = seq(5,45, by = 5))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank(),
        legend.position = "bottom",
        text = element_text(size = 12),
        panel.grid = )+
  ylab("Percent molecules with mean label rating > 1")+
  scale_y_continuous(sec.axis = dup_axis())+
  ggtitle("Relative frequency of labels among contaminated molecules\nand full test set")
```





